An Efficient Parameter Space Search as an Alternative to Markov Chain Monte Carlo

### 2014-07-20 aps4p5/drafts/draft_140719.tex
more writing draft

If the calculation going from $\vec{\theta}\rightarrow\chi^2$ is rapid, this
problem is trivial.  One generates a fine grid of points in the 
$N_p$-dimensional parameter space, evaluates $\chi^2$ at each of these points, 
and defines the confidence limit as only those points which satisfy the rule.  
Often, however, the calcuation $\vec{\theta}\rightarrow\chi^2$ is not rapid, or $N_p$
is large enough that such a grid-based method will take a prohibitively long
time.  In that case, one must make intelligent choices regarding which points in
parameter space are actually interesting enough to warrant
 $\vec{\theta}\rightarrow\chi^2$ evalutation.

@@

\begin{itemize}
\item(1A) Generate some initial number $N_s$ of randomly-distributed points in 
parameter space.  Find the values of $\chi^2$ corresponding to these points.
\\
\item(2A) Generate some number $N_c$ of candidate points.  Use the points
already sampled by APS to predict the value of $\chi^2$ at each of these points.
 We will signify this prediction by $\mu$.  Calculate some value $\sigma$
 quantifying confidence in $\mu$ as a prediction of $\chi^2$.  Low values of
 $\sigma$ should correspond to confident predictions.
 \\
\item(3A) Choose the candidate point which maximizes the statistic
\begin{equation}
\label{eq:sstat}
S=\sigma-|\mu-\chi^2_\text{lim}|
\end{equation}
and evaluate $\vec{\theta}\rightarrow\chi^2$ at that point.  Add this point and
its $\chi^2$ value to the list of sampled points begun in (1A). 
Return to (2A) and repeat until convergence.
\\
\end{itemize}
The confidence limit reported by APS is the list of all points found
for which $\chi^2\le\chi^2_\text{lim}$.  
These points, when plotted in one- or two-dimensional slices of parameter
space, ought to sketch out a region which contains the true value of $\vec{\theta}$
with the desired confidence (as defined in Appendix \ref{sec:frequentism}).
This is in contrast to the credible
limits of MCMC, which represent an integral over the sampled points containing some
fraction of the total posterior probability.

The $|\mu-\chi^2_\text{lim}|$ term in equation (\ref{eq:sstat}) provides an
incentive for APS to choose points that it believes will lie on the boundary of
the confidence limit.  
The $\sigma$ term provides an incentive for APS to explore unknown regions of
parameter space, making the algorithm robust against multimodal $\chi^2$
functions.

Bryan {\it et al.} (2007) show that this algorithm is robust against multi-modal
likelihood functions, using it to identify a disjoint region of high likelihood in
cosmological parameter space as constrained by the 1-year WMAP CMB anisotropy power
spectrum.  This is the principal advantage to selecting points for evaluation according
to the statistic (\ref{eq:sstat}).  Unfortunately, it is possible that the focus on exploring
unknown regions of parameter space might slow down convergence in the case of uni-modal
likelihood functions.  Therefore, we propose the following modifications to the algorithm
to ensure that it adequately characterizes known regions of high likelihood while
simultaneously searching for new regions to explore.

\subsubsection{Characterization of known high likelihood regions}
\label{sec:focus}

As the algorithm searches, it will keep track of a list of all of the discovered points
which are believed to be local minima of $\chi^2$ (how it finds these will be described below
in Section \ref{sec:simplex}).  After each cycle of the main algorithm (steps 1A-3A above),
our version of APS attempts ``focused searches'' about these local minima.

For each local minimum $\vec{M}$, APS performs the following:
\begin{itemize}
\item (1B) If fewer than $N_p$ focused searches have been proposed around this minimum, randomly
choose a point from a small sphere surrounding $\vec{M}$ in parameter space and 
evaluate $\chi^2$ at that point.  
\\
\item(2B) If the point does not satisfy $\chi^2=\chi^2_\text{lim}$ (which is highly probable), 
use bisection
along the line connecting the proposed point to $\vec{M}$ to find a point which does
satisfy $\chi^2=\chi^2_\text{lim}$, i.e. find a point on that line with
$\chi^2<\chi^2_\text{lim}$ -- often $\vec{M}$ itself -- and a point along that line
with $\chi^2>\chi^2_\text{lim}$ and iteratively step half the distance between them until
arriving at the desired $\chi^2_\text{lim}$ \cite{minuit}.  The points discovered this way with
$\chi^2=\chi^2_\text{lim}$ are stored as ``boundary points'' about $\vec{M}$.  Each local
minimum will have its own list of boundary points.
\\
\item(3B) If $N_p$ boundary points have been found for $\vec{M}$, skip steps (1B) and (2B).
Instead, examine the boundary points associated with $\vec{M}$ and identify the points
corresponding to both the maximum and minimum values of each parameter $\theta_i$.  These
points will be referred to as $\vec{B}^{(j)}$.  There will be $2\times N_p$ of them, one for
the maximum and one for the minimum in each of our $N_p$ parameters.  
\\
\item(4B) For each point
$\vec{B}^{(j)}$, propose some number $N_v$ of new points (we have chosen
$N_v=20$ for the sake of speed) $\vec{d}^{(k)} = \vec{B}^{(j)}+\vec{v}$ where $\vec{v}$ is a
random, small vector in parameter space constrained to be perpendicular to
$\vec{B}^{(j)}-\vec{M}$.  There will now be $2\times N_p \times N_v$ points $\vec{d}^{(k)}$
($N_v$ such points for each of the $2\times N_p$ extremal boundary points $\vec{B}^{(j)}$).
\\
\item(5B) At each point $\vec{d}^{(k)}$, evaluate equation (\ref{eq:sstat}).  Evaluate $\chi^2$
at the point $\vec{d}^{(k)}$ which maximizes $S$. If
$\chi^2 \neq \chi^2_\text{lim}$, bisection is again used and a new boundary point is found.
\end{itemize}

We are indebted to Eric Linder of Lawrence Berkeley National Lab who first suggested to us
both the use of bisection to isolate $\chi^2_\text{lim}$ and the idea of stepping
intentionally along the $\chi^2=\chi^2_\text{lim}$ surface by stepping perpendicularly to the
direction $\vec{B}^{(j)}-\vec{M}$.

To reiterate, our version of APS alternates between performing steps (1A-3A) and steps
(1B-5B).  If there are $N_m$ distinct known local minima in $\chi^2$, or algoritm will perform
$N_m$ iterations of steps (1A-3A), and then perform steps (1B-5B) once on each local minimum.

\subsubsection{Extending the boundary}
\label{sec:unitSphere}

The modification suggested in section \ref{sec:focus} is designed to explore 
$\chi^2=\chi^2_\text{lim}$ surface by starting from its extreme corners and stepping
perpendicularly to its ``radii'' (the vector $\vec{B}^{(j)}-\vec{M}$ would be a radius of
$\chi^2=\chi^2_\text{lim}$ described a sphere).

something about APS wide helping to discover new corners of the surface as well

@@

If the user does not feel comfortable settig $\chi^2_\text{lim}$
a priori, APS $\chi^2_\text{lim}$ can learn $\chi^2_\text{lim}$ by
defining it as $\chi^2_\text{min}+\Delta\chi^2$.  Here, $\chi^2_\text{min}$
is the minimum value of $\chi^2$ discovered by APS.  $\Delta \chi^2$ is set
based on the properties of the $N_p$-dimensional parameter space being explored.
To wit, Wilks (1938) proves that $\Delta\chi^2$ for an $N_p$-dimensional
parameter space will be distributed according to a $\chi^2$-distribution
with $N_p$ degrees of freedom.  One can perform the same integral discussed
above and set $\Delta\chi^2$ to that value which encloses the desired confidence
limit.  This is the well-known Likelihood Ratio Test \cite{np}.
In this case, the user is relying on APS to eventually find the true
$\chi^2_\text{min}$ of the likelihood surface being explored.  We show in
Section \ref{sec:wmap} that, with the inclusion of the gradient descent search
discussed above, this is a safe assumption to make.

\subsection{User-specified Parameters of APS}
\label{sec:user}

We make code implementing APS for an arbitrary $\vec{\theta}\rightarrow\chi^2$
function available at \verb|https://github.com/uwssg/APS/|.  
Below, we present a list of user-specified parameters 
required to run the algorithm.  Documentation is also provided with
the code.

\begin{itemize}
\item $N_s$ -- the number of randomly selected points selected in step (1A).
\\

\item $\chi^2_\text{lim}$ -- the target threshold value of $\chi^2$.  Alternatively,
the user can specify
\\
\item $\Delta \chi^2$ -- a threshold such that APS sets
$\chi^2_\text{lim}=\chi^2_\text{min}+\Delta\chi^2$ where $\chi^2_\text{min}$ is the
minimum value of $\chi^2$ discovered by APS at any given time.  This is the framework
adopted in this paper.
\\

\item $N_c$ -- the number of candidate points considered in setp (2A).  Recall
that the code will only evaluate $\vec{\theta}\rightarrow\chi^2$ on the
candidate point which maximizes equation (\ref{eq:sstat}).
\\

\item $N_g$ -- in order to speed up step (2A), the code
as presented only uses the $N_g$ nearest neighbors (in parameter space) of
each candidate point as inputs to prediction $\mu(\vec{\theta})$.
\\

\item $R_g$ -- (\verb|grat| in the code) the ratio determining whether a point
will be considered a candidate for initiating a gradient descent search. 
Gradient descent will only be considered for points for which
$$\chi^2-\chi^2_\text{lim}\le R_g\times \chi^2_\text{lim}$$
\\

\item $N_w$ -- the maximum number of gradient descent searches that the code
will allow to go on in parallel with the ``usual'' APS search at one time.  A
gradient descent search is terminated once it has either
found a point which satisfies $\chi^2\le\chi^2_\text{lim}$ or 
has reduced its step
size such that it is effectively no longer moving through parameter space.  
If all $N_w$ gradient descent slots are already full when another gradient descent
candidate is found, the code will add the new
gradient descent only if its present value of $\chi^2$ is closer to
$\chi^2_\text{lim}$ than the farthest active gradient descent.  In that case,
the farthest active gradient descent will be terminated.
\\

\item $\{\text{min}_i,\text{max}_i\}$ -- 
the minimum and maximum values of the $N_p$ parameters over which you want APS
to search.
\end{itemize}

@@

Gaussian processes take noisy, sparse, measurements of an unknown function (in our
case, $\vec{\theta}\rightarrow\chi^2$ at
the points in parameter space already sampled by APS) and use them to
make inferences about unmeasured values of the function by assuming that the
function is a random process.  Gaussian processes have already found use in physics
and astronomy constructing non-parametric models of the cosmic expansion
\cite{ericgp}, interpolating the point spread function across 
astronomical images
\cite{psf}, and interpolating models of the non-linear matter
power spectrum from N-body simulations \cite{habib}.
They are useful in the current context because, not only do they
give robust predictions $\mu$ for the unmeasured values of the
function being modeled, but they also
provide well-motivated prediction uncertainties $\sigma$, a requirement of
equation (\ref{eq:sstat}).  We present below an introduction to the
formalism of Gaussian processes drawn heavily
from Chapter 2 and Appendix A 
of \cite{gp}.

>>>>

If the calculation going from $\vec{\theta}\rightarrow\chi^2$ is rapid, this
problem is trivial.  One generates a fine grid of points in the 
$N_p$-dimensional parameter space, evaluates $\chi^2$ at each of these points, 
and defines the confidence limit as only those points which satisfy the rule.  
Often, however, the calcuation $\vec{\theta}\rightarrow\chi^2$ is not rapid, or $N_p$
is large enough that such a grid-based method will take a prohibitively long
time.  In that case, one must make intelligent choices regarding which points in
parameter space are actually interesting enough to warrant
 $\vec{\theta}\rightarrow\chi^2$ evaluation.

@@

\begin{itemize}
\item(1A) Generate some initial number $N_s$ of randomly-distributed points in 
parameter space.  Find the values of $\chi^2$ corresponding to these points.
\\
\item(2A) Generate some number $N_c$ of candidate points.  
For each of these points, find the $N_g$ nearest points already sampled
by APS. Use these nearest neighbor points and their corresponding $\chi^2$ values
to predict the value of $\chi^2$ at each of these points.
 We will signify this prediction by $\mu$.  Calculate some value $\sigma$
 quantifying confidence in $\mu$ as a prediction of $\chi^2$.  Low values of
 $\sigma$ should correspond to confident predictions.
 \\
\item(3A) Choose the candidate point which maximizes the statistic
\begin{equation}
\label{eqn:sstat}
S=\sigma-|\mu-\chi^2_\text{lim}|
\end{equation}
and evaluate $\vec{\theta}\rightarrow\chi^2$ at that point.  Add this point and
its $\chi^2$ value to the list of sampled points begun in (1A). 
Return to (2A) and repeat until convergence.
\\
\end{itemize}
The confidence limit reported by APS is the list of all points found
for which $\chi^2\le\chi^2_\text{lim}$.  
These points, when plotted in one- or two-dimensional slices of parameter
space, ought to sketch out a region which contains the true value of $\vec{\theta}$
with the desired confidence (as defined in Appendix \ref{sec:frequentism}).
This is in contrast to the credible
limits of MCMC, which represent an integral over the sampled points containing some
fraction of the total posterior probability.

The $|\mu-\chi^2_\text{lim}|$ term in equation (\ref{eqn:sstat}) provides an
incentive for APS to choose points that it believes will lie on the boundary of
the confidence limit.  
The $\sigma$ term provides an incentive for APS to explore unknown regions of
parameter space, making the algorithm robust against multimodal $\chi^2$
functions.

In the present work, we introduce modifications and expansions to this basic algorithm as
described below.

@@

If the user does not feel comfortable settig $\chi^2_\text{lim}$
a priori, APS $\chi^2_\text{lim}$ can learn $\chi^2_\text{lim}$ by
defining it as $\chi^2_\text{min}+\Delta\chi^2$.  Here, $\chi^2_\text{min}$
is the minimum value of $\chi^2$ discovered by APS.  
$\Delta \chi^2$ is set
based on the properties of the $N_p$-dimensional parameter space being explored.
To wit, Wilks (1938) proves that $\Delta\chi^2$ for an $N_p$-dimensional
parameter space will be distributed according to a $\chi^2$-distribution
with $N_p$ degrees of freedom.  One can perform the same integral discussed
above and set $\Delta\chi^2$ to that value which encloses the desired confidence
limit.  This is the well-known Likelihood Ratio Test \cite{np}.
In this case, the user is relying on APS to eventually find the true
$\chi^2_\text{min}$ of the likelihood surface being explored.  
Section \ref{sec:simplex} below outlines an additional means of search
by which our APS is capable of efficiently locating $\chi^2_\text{lim}$.

\subsection{Modified search algorithms}

Bryan {\it et al.} (2007) show that the APS search algorithm as presented in steps (1A-3A)
is robust against multi-modal
likelihood functions, using it to identify a disjoint region of high likelihood in
cosmological parameter space as constrained by the 1-year WMAP CMB anisotropy power
spectrum.  This is the principal advantage to selecting points for evaluation according
to the statistic (\ref{eqn:sstat}).  Unfortunately, it is possible that the focus on 
exploring
unknown regions of parameter space might slow down convergence in the case of uni-modal
likelihood functions.  Therefore, we propose the following modified search
to ensure that it adequately characterizes known regions of high likelihood while
simultaneously searching for new regions to explore.

These modified searches will require APS to keep track of information it learns about the
$\vec{\theta}\rightarrow\chi^2$ fuction as it searches.  
To facilitate this discussion, we introduce the folling notations.
Symbols denoted with $\vec{x}$ represent points in parameter space.  Variables denoted with
$\{\hat{X}\}$ denote sets of points in parameter space.  Variables denoted $\{X\}$ are sets
of numbers stored by APS.

We are greatly indebted to Eric Linder of the Lawrence Berkeley National
Laboratory who first suggested to us the modifications described below
in sections \ref{sec:focus} and \ref{sec:simplex}.

\subsubsection{Characterization of known low-$\chi^2$ regions}
\label{sec:focus}

As it searches, APS will keep track of a list of all of the discovered points
which are believed to be local minima of $\chi^2$ 
(how it finds these will be described below
in Section \ref{sec:simplex}).  These points will be stored in the set
$\{\hat{\Gamma}\}$.  After each cycle of the main algorithm (steps 1A-3A above),
our version of APS attempts ``focused searches'' about the points in $\{\hat{\Gamma}\}$.

For each point $\vec{M}$ in $\{\hat{\Gamma}\}$, APS performs the following:
\begin{itemize}
\item (1B) If fewer than $N_p$ focused searches have been proposed around this minimum, 
randomly choose a point from a small sphere surrounding $\vec{M}$ in parameter space and 
evaluate $\chi^2$ at that point.  
\\
\item(2B) If the point does not satisfy $\chi^2=\chi^2_\text{lim}$ (which is highly probable), 
use bisection
along the line connecting the proposed point to $\vec{M}$ to find a point which does
satisfy $\chi^2=\chi^2_\text{lim}$, i.e. find a point on that line with
$\chi^2<\chi^2_\text{lim}$ -- often $\vec{M}$ itself -- and a point along that line
with $\chi^2>\chi^2_\text{lim}$ and iteratively step half the distance between them until
arriving at the desired $\chi^2_\text{lim}$ \cite{minuit}.  The points discovered this way \
with $\chi^2=\chi^2_\text{lim}$ are stored in the set $\{\hat{B}(\vec{M})\}$ of 
``boundary points'' about $\vec{M}$ (each $\vec{M}$ in $\{\hat{\Gamma}\}$ will have its own
set $\{\hat{B}(\vec{M})\}$).  
\\
\item(3B) If $N_p$ boundary points have been found for $\vec{M}$, skip steps (1B) and (2B).
Instead, examine the points in $\{\hat{B}(\vec{M})\}$ and identify the points
corresponding to both the maximum and minimum values of each parameter $\theta_i$.  These
points will be referred to as $\{\hat{B}_\text{MaxMin}(\vec{M})\}$.  
There will be $2\times N_p$ points in $\{\hat{B}_\text{MaxMin}(\vec{M})\}$, one for
the maximum and one for the minimum in each of our $N_p$ parameters.  
\\
\item(4B) For each point $\vec{b}$ in $\{\hat{B}_\text{MaxMin}(\vec{M})\}$, 
propose some number $N_v$ of new points (we have chosen
$N_v=20$ for the sake of speed) $\vec{v} = \vec{b}+\vec{\epsilon}$ where $\vec{\epsilon}$ is a
random, small vector in parameter space constrained to be perpendicular to
$\vec{b}-\vec{M}$.  There will now be $2\times N_p \times N_v$ points $\vec{v}$
($N_v$ such points for each of the $2\times N_p$ extremal boundary points in
$\{\hat{B}_\text{MaxMin}(\vec{M})\}$).
\\
\item(5B) At each point $\vec{v}$ above, evaluate equation (\ref{eqn:sstat}).  
Evaluate $\chi^2$ at the point $\vec{v}$ which maximizes $S$. If
$\chi^2 \neq \chi^2_\text{lim}$, bisection is again used and a new boundary point is found.
\end{itemize}

To reiterate, our version of APS alternates between performing steps (1A-3A) and steps
(1B-5B).  If there are $N_m$ distinct known local minima in $\{\hat{\Gamma}\}$, 
our algoritm will perform
$N_m$ iterations of steps (1A-3A), and then perform steps (1B-5B) once for each
$\vec{M}$ in $\{\hat{\Gamma}\}$.

\subsubsection{Extension of the $\chi^2=\chi^2_\text{lim}$ contour}
\label{sec:unitSphere}

The modification suggested in section \ref{sec:focus} is designed to explore 
$\chi^2=\chi^2_\text{lim}$ surface by starting from its extreme corners and stepping
perpendicularly to its ``radii'' (the vector $\vec{b}-\vec{M}$ would be a radius of
$\chi^2=\chi^2_\text{lim}$ described a sphere).  It is, however, also possible
that the $S$-maximization in steps (1A-3A) could also learn something about the
shape of the $\chi^2=\chi^2_\text{lim}$ contour, even if it does not place
points directly on that contour.  To explore that possibility, we add the
following steps to the end of the sequence (1A-3A).

\begin{itemize}
\item(1C) For each point $\vec{M}$ in $\{\hat{\Gamma}\}$, store, 
in addition to the boundary
points $\{\hat{B}(\vec{M})\}$, the projection of those boundary points
onto a unit sphere in parameter space, i.e. for each point $\vec{b}$ in $\{\hat{B}(\vec{M})\}$, draw a line
connecting $\vec{b}$ with $\vec{M}$ and keep track of
where a unit sphere centered on $\vec{M}$ intersects that line.  Call this set
of projectsion $\{\hat{U}\}$.
\\
\item(2C) Call the point for which $\chi^2$ has been evaluated in step (3A)
$\vec{P}$., find the
point $\vec{M}$ in $\{\hat{\Gamma}\}$ that is nearest to $\vec{P}$ in parameter space.
\\
\item(3C) Find the
projection of $\vec{P}$ onto the unit sphere surrounding $\vec{M}$.  
Call this projected point $\vec{p}$.  Find the point $\vec{u}$ in
$\{\hat{U}\}$ that is nearest to $\vec{p}$.  Record the parameter space 
distance between $\vec{u}$ and $\vec{p}$.
Add this distance to the set $\{D\}$ ($\{D\}$ will be a running list
of the $\vec{u}-\vec{p}$ distance for all points $\vec{P}$ found by steps
1A-3A, regardless of which point in $\vec{M}$ they are nearest).
\\
\item(4C) If the $\vec{p}-\vec{u}$ distance is greater than the 2/3 quantile of $\{D\}$,
then perform bisection between $\vec{M}$ and the point evaluated in step (3A).
\end{itemize}

This way, points on the $\chi^2=\chi^2_\text{lim}$ contour can be found whose
directions away from $\vec{M}$ in parameter space have not yet been explored.
Because we expect the unit spheres surrounding the points $\vec{M}$ to become more densely
populated as the algorithm runs, we periodically delete the list $\{D\}$ so that
new directions away from $\vec{M}$ can still be explored.

The final modification to the original APS algorithm which we propose in this
work is to utilize function minimization to identify local minima in $\chi^2$.
One of the principal drawbacks of the algorithm presented in steps (1A-3A) is
that it makes no account of ``near misses.''  If step (3A) finds that $\chi^2$
is lower than average but still greater than $\chi^2_\text{lim}$, the original
APS algorithm does nothing with that information.  We have tried to improve upon
that with our incorporation of bisection in sections \ref{sec:focus} and
\ref{sec:unitSphere} above.  Our final innovation is to allow APS to use points
with 
\begin{eqnarray}
\chi^2_\text{lim}&\le&\chi^2\le\chi^2_\text{threshold}\label{eqn:simplexCriterion}\\
\chi_\text{threshold}&=&\chi^2_\text{min}+G\times(\chi^2_{1/10}-\chi^2_\text{min})
\label{eqn:threshold}
\end{eqnarray}
to seed a simplex function minimizer as developed by Nelder and Mead (1965).
Here, $G$ is a user-set parameter, $\chi^2_\text{min}$ is the minimum discovered
$\chi^2$ value and $\chi^2_{1/10}$ is the 1/10 quantile of the $\chi^2$ values
discovered by steps (1A-3A) (not those discovered by any of the modifications
discussed in sections \ref{sec:focus}, \ref{sec:unitSphere} or
\ref{sec:simplex}).

Because the Nelder-Mead simplex minimizer can be very time intensive, we balance our
algorithm by demanding that the number of $\chi^2$ evaluations devoted to all
other variants of search (i.e. steps (1A-3A) and (1B-5B)) equal the number of
$\chi^2$ evaluations devoted to the simplex minimizer before a new simplex
minimization can occur.

If the simplex minimizer converges to a point such that
$\chi^2<\chi^2_\text{lim}$, our APS determines whether or not it
has found a new local minimum to be added to $\{\hat{\Gamma}\}$ by comparing the end
point $\vec{y}$ of the minimization to each of the points already in
$\{\hat{\Gamma}\}$.  For each $\vec{M}$ in $\{\hat{\Gamma}\}$, APS calculates the mid-point
\begin{equation}
\label{eqn:midpt}
\vec{m}=0.5\times\left(\vec{y}+\vec{M}\right)
\end{equation}
and $\vec{M}$.  If, for all $\vec{M}$, 
$\chi^2$ at $\vec{m}$ is greater
than $\chi^2_\text{lim}$, then it is deemed that $\vec{\gamma}$ is a new local
minimum and is added to $\{\hat{\Gamma}\}$.  If not, $\vec{\gamma}$ is assumed to be a
part of the one of the previously identified loci of low $\chi^2$ and it is not
added to $\{\hat{\Gamma}\}$.  Regardless of whether or not $\vec{y}$ passes the test in
equation (\ref{eqn:midpt}), $\vec{y}$ is added to the set $\{\hat{Y}\}$ of all
simplex end points discovered by APS.

As a further guard against excessive simplex searching, all points $\vec{w}$ 
that pass the test in equation(\ref{eqn:simplexCriterion}) are further required
to satisfy the condition
\begin{eqnarray}
\chi^2(\vec{z})&\le&\chi^2(\vec{w})-0.25\times\left(\chi^2(\vec{w})-\chi^2(\vec{y})\right)\\
\vec{z}&\equiv&0.5\times\left(\vec{w}+\vec{y}\right)\\
\vec{y}&\in&\{\hat{Y}\}
\end{eqnarray}
before it is allowed to seed a new simplex search.  In this way, our APS avoids seeding simplex
searches with points that are inside of already known regions of low $\chi^2$.

\subsection{Code implementing APS}
\label{sec:user}

We make code implementing our extended APS for 
an arbitrary $\vec{\theta}\rightarrow\chi^2$
function available at \verb|https://github.com/uwssg/APS/|.
The code is written in C++.  It provides the option for several
user-specified parameters.  We list them below along with the C++ 
methods (belonging to the class \verb|aps|) used to set them.  
This code includes several user-specified parameters which we
outline below.

\begin{itemize}
\item The parameter $G$ from equation (\ref{eqn:threshold}) above can be set using the method
\verb|aps.set_grat|
\\
\item $\chi^2_\text{lim}$ can be manually set using the method \verb|aps.set_target|
\\
\item $\Delta \chi^2$ can be set in the constructor for the class \verb|aps|.
\\
\item The number $N_g$ of nearest neighbors used in step 2A can be set in the constructor
for the class \verb|aps|
\\
\item $N_c$ from step 2A can be set using the method \verb|aps.set_samples|
\\
\item The minimum and maximum bounds in parameter space can be set using the 
method \verb|aps.set_max| and \verb|aps.set_min|.  These bounds will constrain the range of
points sampled in step 2A.  It will not contraint searches performed with bisection or the 
simplex minimizer.
\\
\item If there is a characteristic length scale associated with a parameter, this can be set
using the method \verb|aps.set_characteristic_length|.  This length scale will be used to
normalize all distances in parameter space.  If it is not set, it will default to the
difference between the maximum and minimum values set by \verb|aps.set_max| and
\verb|aps.set_min|.
\end{itemize}

More complete documentation is made available with the source code.

\subsection{Interpreting APS outputs}
\label{sec:bayes}

Originally, APS was conceived as a way to draw Frequentist confidence limits. 
$\chi^2_\text{lim}$ is determined, either in a vacuum or as some $\Delta\chi^2$ away from
$\chi^2_\text{min}$ (see Appendix \ref{sec:frequentism} for a more thorough discussion).
This is by far the most direct way to interpret APS outputs: a list of points is given with
corresponding values of $\chi^2$.  All of the points with $\chi^2\le\chi^2_\text{lim}$ are
taken to circumscribe the desired confidence limit.  There is, however, also a way to use the
outputs from APS to draw a Bayesian credible limit.

As discussed in the introduction, Bayesian parameter constraints assume that there is some
probability distribution (``the posterior'') over $\vec{\theta}$.  
$(1-\alpha)\%$ credible limits are drawn by
integrating over this probability distribution and selecting that region which contains
$(1-\alpha)\%$ of the total probability.  This integral can be very time-consuming if the
dimensionality of $\vec{\theta}$ is large.  Markov Chain Monte Carlo methods overcome this
difficulty by randomly drawing samples from the posterior and assuming that, if enough samples
are drawn, the distribution of those samples will be sufficient to reconstruct the attributes
of the full posterior probability density.

APS outputs cannot be used this way, since APS does not select points to evaluate
probabilisticaly, but in an attempt to recreate the behavior of the $\chi^2$ function in
interesting regions of parameter space.  We can still, however, use these points to approximate
the integral over the posterior probability density.  The algorithm we propose is as follows.
The total set of points sampled by APS will be referred to below as $\{\hat{A}\}$.

\begin{itemize}
\item(1D) For each point $\vec{a}$ in $\{\hat{A}\}$ initialize two vectors $\vec{a}_\text{max}$ and
$\vec{a}_\text{min}$.

\item(2D) Select the $3N_p+1$ nearest neighbor points to $\vec{a}$ from
$\{\hat{A}\}$ (this will obviously include $\vec{a}$ itself, which can be discarded).  Call
this set of nearest neighbors $\{\hat{N}\}$.
\\
\item(3D) For each point $\vec{n}$ in $\{\hat{N}\}$, calculate the vector
$\vec{\delta}=\|\vec{n}-\vec{a}|$ (where the vertical bars denote the absolute value of the
components of $\vec{\delta}$, not any kind of vector normalization).  For each component
$\delta_i$ of $\vec{\delta}$, starting with the largest and working towards the smallest:
\\
\begin{itemize}
\item(3Da) If $n_i>a_i$ and $a_{\text{max},i}$ has not been set, set $a_{\text{max},i}=n_i$.
\\
\item(3Db) If $n_i<a_i$ and $a_{\text{min},i}$ has not been set, set $a_{\text{min},i}=n_i$.
\\
\item(3Dc) If both $a_\text{max,i}$ and $a_\text{min,i}$ have already been set, move on to the
next largest component of $\vec{\delta}$ and repeat at step (3Da).
\end{itemize}
\item(4D) Choose the next neighbor $\vec{n}$ and repeat from step (3D). Continue until all
$a_{\text{max},i}$ and $a_{\text{min},i}$ have been set.  Once all $a_{\text{max},i}$ and
$a_{\text{min},i}$ have been set, move on to step (5D).
\\
\item(5D) $\vec{a}_\text{max}$ an $\vec{a}_\text{min}$ should now describe an asymmetric,
rectangular hyperbox about $\vec{a}$.  To symmetrize it, step through all $N_p$ dimensions
and reset either $a_{\text{max},i}$ or $a_{\text{min},i}$ so that 
$d_\text{min}=a_i-a_{\text{min},i}$ and
$d_\text{max}=a_{\text{max},i}-a_i$ are equal to the minimum of $\{d_\text{max},d_\text{min}\}$
from the asymmetric hyperbox.
\end{itemize}

At this point, each $\vec{a}$ in $\{\hat{A}\}$ should have a corresponding symmetric hyperbox
surrounding it.  To integrate the posterior probability density, we will assume that all of the
parameter space points in each hyperbox correspond to a $\chi^2$ equal to $\chi^2(\vec{a})$.
The posterior probability associated with each box is then
\begin{eqnarray}
P(\vec{a})&=&\frac{V(\vec{a})}{P_\text{tot}}\times\exp\left[-\chi^2(\vec{a})/2\right]
\label{eqn:posterior}\\
P_\text{tot}&\equiv&\sum_{\vec{a}} V(\vec{a})\times\exp\left[-\chi^2(\vec{a})/2\right]
\end{eqnarray}
where $V(\vec{a})$ is the hypervolume associated with the hyperbox surrouning $\vec{a}$.
To plot the $(1-\alpha)\%$ credible limit, one need only plot the points in hyperspace that
enclose the lowest-$\chi^2$ hyperboxes containing $(1-\alpha)\%$ of the total probability.
We test this algorithm on a function with an exactly known credible limit
in Section \ref{sec:toy} and on a real, physical dataset in Section \ref{sec:wmap}.  In 
Section \ref{sec:toy}, we find that our results agree with the known credible limit.
In Section \ref{sec:wmap}, we find that our results agree with results derived from traditional
Markov Chain Monte Carlo methods.

The functionality described by this section is provided in by the classes defined in the source
code file \verb|aps_extractor.cpp| in our software package.

@@

Gaussian processes take noisy, sparse, measurements of an unknown function (in our
case, $\vec{\theta}\rightarrow\chi^2$ at
the points in parameter space already sampled by APS) and use them to
make inferences about unmeasured values of the function by assuming that the
function is a random process.  Gaussian processes have already found use in physics
and astronomy constructing non-parametric models of the cosmic expansion
\cite{ericgp}, interpolating the point spread function across 
astronomical images
\cite{psf}, and interpolating models of the non-linear matter
power spectrum from N-body simulations \cite{habib}.
They are useful in the current context because, not only do they
give robust predictions $\mu$ for the unmeasured values of the
function being modeled, but they also
provide well-motivated prediction uncertainties $\sigma$, a requirement of
equation (\ref{eqn:sstat}).  We present below an introduction to the
formalism of Gaussian processes drawn heavily
from Chapter 2 and Appendix A 
of \cite{gp}.

Gaussian processes use the sampled data $\{\vec{\theta}^{(i)},\chi^{2,(i)}\}$, 
where $i$ indexes over all of the sampled points, to predict
$\mu=\chi^{2,(q)}$ at the candidate point 
$\vec{\theta}^{(q)}$ by assuming that the function
$\vec{\theta}\rightarrow\chi^2$ represents a sample drawn from a random process 
distributed
across parameter space.  At each point in parameter space, 
$\chi^2$ is assumed to be
distributed according to a normal distribution with mean 
$\bar{\chi}^2$
and variance dictated by the covariance function 
$C_{ij}(\vec{\theta}^{(i)},\vec{\theta}^{(j)})$.
If we have $N_g$ measurements of $\chi^2$, then we assume
they are distributed according to
\begin{equation}
\label{eqn:likelihood}
P(\vec{\chi}^2)=
\frac{\exp\bigg[-\frac{1}{2}K^{-1}\sum_{i,j}^{N_g}
(\chi^{2,(i)}-\bar{\chi}^2)C^{-1}_{ij}
(\chi^{2,(j)}-\bar{\chi}^2)\bigg]}{\sqrt{2\pi}^{N_g}\text{det}|KC|^{N_g/2}}
\end{equation}
$C_{ij}$ is a function assumed by the user encoding how variations in
$\chi^2$ at one point in parameter space affect variations in 
$\chi^2$ at other
points in parameter space.  $K$ is a parameter controlling the
normalization of $C_{ij}$.
The diagonal elements $C_{ii}=1+\sigma^2_{ii}$ where 
$\sigma^2_{ii}$ is the intrinsic variance in the value
of $\chi^2$ at a single point $\vec{\theta}^{(i)}$.  
Rasmussen and Williams (2006) treat 
the special case
$\bar{\chi}^2=0$ and find (see their equations 2.19, 2.25 and 2.26)
\begin{eqnarray}
\mu&=&\sum_{i,j} C_{qi}C^{-1}_{ij}\chi^{2,(j)}\nonumber\\
\sigma^2&=&C_{qq}-\sum_{i,j}C_{qi}C^{-1}_{ij}C_{jq}\label{sig}
\end{eqnarray}
where the sums over $i$ and $j$ are sums over the sampled points 
in parameter space.  $C_{iq}$ relates the sampled point $i$ to the candidate
point $q$.
We do not wish to assume that the mean value of $\chi^2$ is zero everywhere.
Therefore, we modify the equation for $\mu$ to give
\begin{equation}
\label{mu}
\mu=\bar{\chi}^2+\sum_{i,j} C_{qi}C^{-1}_{ij}
(\chi^{2,(j)}-\bar{\chi}^2)
\end{equation}
where $\bar{\chi}^2$ is the algebraic mean of the 
sampled $\chi^{2,(i)}$.
Note the similarity to a multi-dimensional Taylor series expansion with the
covariance matrix playing the role of the derivatives.
Equation (\ref{sig}) differs from equation (6) in \cite{bryan} because
they used the semivariance
$$\gamma_{ij}=\text{var}[\chi^2(\vec{\theta}^{(i)})
-\chi^2(\vec{\theta}^{(j)})]$$
in place of the covariance $C_{ij}$.  In practice, the two assumptions result in
equivalently valid $\mu$ and $\sigma$.
